# Symmetry and Asymmetry - Plan

Within studies of biface morphology, an analysis of bilateral/object symmetry can yield useful insights related to _design intent_, given the assumption that the knapper's goal was to achieve a symmetrical shape. Within that framework, deviation from bilateral symmetry (fluctuating asymmetry) can serve as a useful measure of _morphological standardization_ within an analysis of morphological disparity. Both measures have utility in comparisons of archaeological contexts and sites, to include wide-ranging spatial and temporal applications.

## Gahagan bifaces

Per Klingenberg [-@RN8579, 1912 and Figure 2]:

* Symmetric variation includes movements of the unpaired landmarks along the median axis (or within the median plane) and joint mirror-image shifts of pairs of corresponding landmarks in any direction.
* Asymmetry is quantified through the differences between the original and mirrored configuration from the symmetric consensus of the original and mirror image.

## Load packages + data

```{r load.data.asymm, echo = TRUE, cache = TRUE}
# load packages
#devtools::install_github("geomorphR/geomorph", ref = "Stable", build_vignettes = TRUE)
library(here)
library(geomorph)
library(tidyverse)
library(wesanderson)

# read GM data
source('readmulti.csv.R')

# read .csv files
setwd("./data.3d1")
filelist <- list.files(pattern = ".csv")
coords <- readmulti.csv(filelist)
setwd("../")

# read qualitative data
qdata <- read.csv("qdata.csv", header = TRUE, row.names = 1)
qdata <- qdata[match(dimnames(coords)[[3]], rownames(qdata)),]
land.pairs <- read.csv("asymmetry.csv", header = TRUE)
ind <- rep(1:length(filelist))
```

## Generalized Procrustes Analysis

Landmark data were aligned to a global coordinate system [@RN11622;@RN11623;@RN11563], achieved through generalized Procrustes superimposition [@RN478] performed in R 4.0.2 [@R] using the `geomorph` library v. 3.3.1 [@RN11530;@RN1774]. Procrustes superimposition translates, scales, and rotates the coordinate data to allow for comparisons among objects [@RN11564;@RN478]. The `geomorph` package uses a partial Procrustes superimposition that projects the aligned specimens into tangent space subsequent to alignment in preparation for the use of multivariate methods that assume linear space [@RN1646;@RN11563].

```{r asymm.gpa, echo=TRUE, warning=FALSE}
# gpa - bilateral asymmetry
res.bilat <- bilat.symmetry(coords,
                          ind = ind, 
                          object.sym = TRUE,
                          land.pairs = land.pairs,
                          RRPP = TRUE,
                          iter = 9999,
                          print.progress = FALSE)
summary(res.bilat)
```

```{r symm.shape, out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE, cache = TRUE, fig.cap="_**Figure 3a-b. Mean shape (black) and specimens (gray) for the symmetric shape component plotted in 2D.**_"}
# symmetric shape/design intent
symm.shape <- res.bilat$symm.shape[,1:2,]
symm.shape2 <- res.bilat$symm.shape[,2:3,]

# plot all specimens/2D (symmetric component)
plotAllSpecimens(symm.shape)
plotAllSpecimens(symm.shape2)
```

```{r asymm.shape, out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE, cache = TRUE, fig.cap="_**Figure 4a-b. Mean shape (black) and specimens (gray) for the asymmetric shape component plotted in 2D.**_"}
# asymmetric shape/standardization?
asymm.shape <- res.bilat$FA.component[,1:2,]
asymm.shape2 <- res.bilat$FA.component[,2:3,]

# plot all specimens/2D (asymmetric component)
plotAllSpecimens(asymm.shape)
plotAllSpecimens(asymm.shape2)
```

## Principal Components Analysis

Principal components analysis [@RN1746] was used to visualize shape variation among the bifaces. The shape changes described by each principal axis are commonly visualized using thin-plate spline warping of a reference 3D mesh [@RN1731;@RN479].

```{r pca.asymm, out.width = "100%", dpi = 300, echo=TRUE, cache = TRUE, warning=FALSE}
# geomorph data frame
gdf <- geomorph.data.frame(shape = res.bilat$symm.shape,
                           ashape = res.bilat$FA.component,
                           cntxt = qdata$context) 

# attributes
cntxt <- qdata$context

# principal components analysis (symmetric)
pca <- gm.prcomp(res.bilat$symm.shape)
summary(pca)

# set plot parameters to plot by burial context
pch.gps.cntxt <- c(3,4,15,17:19)[as.factor(cntxt)]
col.gps.cntxt <- wes_palette("IsleofDogs1")[as.factor(cntxt)]
col.hull.cntxt <- c("#9986A5","#8D8680","#D9D0D3","#CCBA72","#0F0D0E","#79402E")
```

## Define models

### Hypothesis

```{r def.mod.1.asymm, out.width = "100%", dpi = 300, echo=TRUE, cache = TRUE, warning=FALSE}
# symmetric shape as a function of burial context
## (differences in design intent?)
reg.symm <- procD.lm(shape ~ cntxt, 
                     data = gdf, 
                     print.progress = FALSE, 
                     iter = 9999)

# asymmetric shape as a function of burial context
## (standardization?)
reg.asymm <- procD.lm(ashape ~ cntxt, 
                      data = gdf, 
                      print.progress = FALSE, 
                      iter = 9999)
```

```{r diagnostic, out.width = "100%", dpi = 300, echo=FALSE, cache = TRUE, warning=FALSE}
#par(mfrow = c(2,3))
#plot(reg.symm, type = "diagnostic", outliers = TRUE)
```

```{r diagnostic.asymm, out.width = "100%", dpi = 300, echo=FALSE, warning=FALSE, cache = TRUE}
#plot(reg.asymm, type = "diagnostic", outliers = TRUE)
```

## Test Hypothesis

Assessment of _design intent_ and _standardization_ by region.

### PCA (H1)

```{r symm.pca, out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE, cache = TRUE}
# plot pca (symmetric) by burial context
pc.plot <- plot(pca, 
                 asp = 1,
                 pch = pch.gps.cntxt,
                 col = col.gps.cntxt)
                    shapeHulls(pc.plot, 
                               groups = cntxt,
                               group.cols = col.hull.cntxt)
```

### Procrustes ANOVA

_Are there differences in the symmetric shapes of Gahagan bifaces found in the southern Caddo area (north/south) and central Texas?_

A residual randomization permutation procedure (RRPP; n = 10,000 permutations) was used for the Procrustes ANOVA [@RN1655;@RN11775], which has higher statistical power and a greater ability to identify patterns in the data should they be present [@RN1719]. To assess whether symmetric shape differs by group (region), Procrustes ANOVAs [@RN1749] were also run that enlist effect-sizes (zscores) computed as standard deviates of the generated sampling distributions [@RN1756].

```{r h1.asymm, out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE, cache = TRUE}
# ANOVA: do symmetric gahagan biface point shapes differ by region?
## (are there differences in design intent?)
anova(reg.symm)

# pairwise comparison of LS means = which differ?
symm.reg <- pairwise(reg.symm,
                     groups = qdata$context)

summary(symm.reg, 
        confidence = 0.95, 
        test.type = "dist")

# pairwise distance between variances = standardization?
summary(symm.reg, 
        confidence = 0.95, 
        test.type = "var")
```

### Mean Symmetric Shapes

Mean symmetric shapes were rendered and contrasted for each region, and illustrate those morphological differences that occur between each. These symmetric shapes may represent regional differences in _design intent_ related to the conceptual aesthetic of Gahagan bifaces, which differs significantly between some of the burials (see above).

### Morphological disparity

_Does fluctuating asymmetry (standardization) vary between regions?_

Morphological disparity is estimated as the Procrustes variance, for groups in this instance, using residuals of a linear model fit [@RN11530]. Procrustes variance is the same sum of the diagonal elements of the group covariance matrix divided by the number of observations in the group [@RN5703;@RN11530]. Absolute differences in Procrustes variances are test statistics that can be used to test differences in morphological disparity among groups [@RN11530]. These differences are statistically evaluated through permutation, where the vectors of residuals are randomized among groups [@RN11530].

```{r m-disparity.region, out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE, cache = TRUE}
# morphological disparity: do any of the groups display greater asymmetric 
# shape variation among individuals relative to the other group?
## (standardization?)
morphol.disparity(reg.asymm, 
                  groups = qdata$context, 
                  data = gdf, 
                  print.progress = FALSE, 
                  iter = 9999)
```
